---
title: "rnz R NetCDF Zarr"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rnz R NetCDF Zarr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
   .main-container {max-width: 100%;}
   .row {display: flex;}
   .column {flex: 50%;}
</style>

```{r setup, include = FALSE}
eval <-  requireNamespace("RNetCDF", quietly = TRUE) & requireNamespace("pizzarr", quietly = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.width=4,
  fig.height=4,
  fig.align = "center",
  eval=eval
)

oldoption <- options(scipen = 9999)
```

# `rnz` R NetCDF Zarr

`rnz` wraps the `pizzarr` package providing very similar set of functions to the 
`RNetCDF` package. The intent is to allow access to zarr in essentially the same 
way as is done with NetCDF such that packages with a NetCDF backend can easily
implement an equivalent ZARR back end using base R.

This vignette shows the most basic operation of the package using a core
demo dataset derived from: https://gdo-dcp.ucllnl.org/downscaled_cmip_projections/

A NetCDF and Zarr copy of the dataset are included.

```{r}

# these files are in `z_demo()` as a file store
# and via "https://raw.githubusercontent.com/DOI-USGS/rnz/main/inst/extdata/bcsd.zarr/" as an http store.
gsub(normalizePath(dirname(rnz::z_demo())), "",
     list.files(rnz::z_demo(), recursive = TRUE, all.files = TRUE), 
     fixed = TRUE)

(z <- rnz::open_zarr(rnz::z_demo()))

cat(c(capture.output(rnz::zdump(rnz::z_demo()))[1:25], "..."), sep = "\n")

nc_file <- system.file("extdata", "bcsd_obs_1999.nc", package = "rnz")

(nc <- RNetCDF::open.nc(nc_file))

cat(c(capture.output(RNetCDF::print.nc(nc))[1:25], "..."), sep = "\n")

```

## "Inquire" about elements of a Zarr store

We can do all the normal inquire stuff using 0-indexed ids or names.

### Inquire about a store 

```{r}
RNetCDF::file.inq.nc(nc)

rnz::inq_store(z)
```

### Inquire about a group

TODO: get rnz working with more than root groups

```{r}

RNetCDF::grp.inq.nc(nc)

rnz::inq_grp(z)

```

### Inquire about a dimension

```{r}

RNetCDF::dim.inq.nc(nc, 2)

rnz::inq_dim(z, 2)

RNetCDF::dim.inq.nc(nc, "time")

rnz::inq_dim(z, "time")
```

### Inquire about a variable

```{r}

RNetCDF::var.inq.nc(nc, 4)

rnz::inq_var(z, 4)

RNetCDF::var.inq.nc(nc, "time")

rnz::inq_var(z, "time")
```

### Inquire about an attribute

```{r}
RNetCDF::att.inq.nc(nc, "time", "units")

rnz::inq_att(z, "time", "units")

RNetCDF::att.inq.nc(nc, 4, 1)

rnz::inq_att(z, 4, 3)
```
### Get a variable

```{r}

RNetCDF::var.get.nc(nc, "time")

rnz::get_var(z, "time")

RNetCDF::var.get.nc(nc, variable = "pr", 
                    start = c(1,1,5), count = c(3,3,1))

rnz::get_var(z, var = "pr", 
             start = c(1,1,5), count = c(3,3,1))

pr <- rnz::get_var(z, "pr") |>
  aperm(c(3,2,1))

# TODO: figure out NA handling
pr[pr > 1000] <- NA

image(pr[,,1], col = hcl.colors(12, "PuBuGn", rev = TRUE),
      useRaster = TRUE, axes = FALSE)
```

```{r}
pr <- RNetCDF::var.get.nc(nc, "pr")

image(pr[,,1], col = hcl.colors(12, "PuBuGn", rev = TRUE),
      useRaster = TRUE, axes = FALSE)
```

### Get an attribute

```{r}
RNetCDF::att.get.nc(nc, "time", "units")

rnz::get_att(z, "time", "units")

RNetCDF::att.get.nc(nc, 4, 1)

rnz::get_att(z, 4, 3)
```
That's all for now.

```{r include = FALSE}

options(oldoption)

```
